---
title: "Assignment 1 - Chapter 8-19 Summary LK"
author: "Lynnstacy Kegeshi"
date: "2024-10-28"
output: 
        pdf_document: 
          toc: yes
        html_document:
          toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This is a summary of Chapter 8-19 of the book **R for Data Science**. We shall be using be data set *House Sale* to perform our analysis based on the content from Chapter 8 - 19

We shall be using the following libraries
```{r, echo=TRUE, results='hide'}
library(tidyverse)
library(stringr)
library(forcats) # Factor manipulation
library(lubridate) # Date and time
library(magrittr) #For Pipe operation
```

# Chapter 8
This chapter focuses on importing data into R, specifically using the read r package which is part of tidyverse.

## Reading Data into R Studio
Most functions in readr are designed to turn flat files into data frames. Key functions include:

- read_csv(): Reads comma-delimited files.
- read_csv2(): Reads semicolon-separated files.
- read_tsv(): Reads tab-delimited files.
- read_delim(): Reads files with any delimiter.
- read_fwf(): Reads fixed-width files.
- read_log(): Reads Apache-style log files.

Using the read_csv(), we imported the HouseSale Data to illustrate this fucntion.

```{r}
house_sale<- read.csv("HouseSale.csv")
head(house_sale)
```

You can also customize the import of data through the following options.

1. Skipping Metadata: Use skip = n to skip the first n lines or comment = "#".
2. No Column Names: Use col_names = FALSE to treat the first row as data.
3. Custom Column Names: Pass a character vector to col_names.
4. Handling Missing Values: Use the na argument to specify missing value representations.

## Writing into a file
`readr also provides functions for writing data back to disk:

- write_csv(): Writes CSV files.
- write_tsv(): Writes tab-separated files.
- write_excel_csv(): Writes CSV files for Excel.
- write_rds() and read_rds(): Store data in R's custom binary format.
- write_feather() and read_feather(): Use the feather package for fast binary file format.

## Parsing Vectors
The parse_*() functions are designed to convert character vectors into more specialized types. Here are some of the most commonly used functions

- parse_logical(): Parses logical values.
- parse_integer(): Parses integer values.
- parse_double(): Parses double (numeric) values.
- parse_number(): Parses numbers, ignoring non-numeric characters.
- parse_character(): Parses character strings.
- parse_factor(): Parses factors.
- parse_datetime(): Parses date-time values.
- parse_date(): Parses date values.
- parse_time(): Parses time values.

```{r}
# Parsing logical values
logical_vector <- parse_logical(c("TRUE", "FALSE", "NA"))
print(logical_vector)
#> [1] TRUE FALSE NA

# Parsing integer values
integer_vector <- parse_integer(c("1", "2", "3"))
print(integer_vector)
#> [1] 1 2 3

# Parsing double values
double_vector <- parse_double(c("1.23", "4.56", "7.89"))
print(double_vector)
#> [1] 1.23 4.56 7.89

# Parsing numbers with non-numeric characters
number_vector <- parse_number(c("$100", "20%", "It cost $123.45"))
print(number_vector)
#> [1] 100  20 123.45

# Parsing date values
date_vector <- parse_date(c("2010-01-01", "1979-10-14"))
print(date_vector)
#> [1] "2010-01-01" "1979-10-14"
```
We check the structure of our data to see if we need to parse any data.

```{r}
str(house_sale)
```
Since our data(House Sale) is already structured with the correct types the parse_* functions are unnecessary here. 

# Chapter 9

This chapter introduces the concept of tidy data, a consistent way to organize data in R. This is done using tidyr which is a package inside tidyverse.

Tidy data is organized according to three interrelated rules:

1. Each variable must have its own column.
2. Each observation must have its own row.
3. Each value must have its own cell.

Tidy data has two main advantages: **Consistency** and **Efficiency**

## Missing Values
Missing values can be explicit (flagged with NA) or implicit (not present in the data). tidyr provides tools to handle missing values:

complete(): Ensures all combinations of variables are present, filling in missing values with NA.
Applying this to our data where StreetHouseFront had explicit missing values, we filled missing values with the median on non missing values.

##Applying ggplot and dplyr on tidydata
```{r}
# Fill missing StreetHouseFront values with the median of non-missing values
house_sale <- house_sale %>%
  mutate(StreetHouseFront2 = ifelse(is.na(StreetHouseFront), median(StreetHouseFront, na.rm = TRUE), StreetHouseFront))

#Chcking if NAs had been replaced.
head(house_sale$StreetHouseFront)
head(house_sale$StreetHouseFront2)
```
fill(): Fills missing values with the most recent non-missing value.

```{r}
# Fill NA values in StreetHouseFront with the most recent non-NA value
house_sale3 <- house_sale %>%
  fill(StreetHouseFront)

#Chcking if NAs had been replaced.
head(house_sale$StreetHouseFront)
head(house_sale3$StreetHouseFront)

```
The chapter also provides examples of working with tidy data using dplyr and ggplot2

```{r}
# Calculate average house price by Location
house_sale %>%
  group_by(Location) %>%
  summarize(avg_price = mean(HousePrice, na.rm = TRUE))#na.rm ignores missing values
```
```{r}
# Calculate average rating by SellingYear
avg_rating_per_year <- house_sale %>%
  group_by(SellingYear) %>%
  summarize(avg_rating = mean(Rating, na.rm = TRUE))

# Plot average rating over time
ggplot(avg_rating_per_year, aes(x = SellingYear, y = avg_rating)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  labs(title = "Average Property Rating Over the Years", x = "Year Sold", y = "Average Rating") +
  theme_minimal()
```

Most real-world data is not tidy. Two common problems are:

- One variable spread across multiple columns.
- One observation scattered across multiple rows.

To address these issues, tidyr provides two key functions: gather() and spread().

'separate() and unite() are used to split and combine columns, respectively.

# Chapter 10
Keys are variables that uniquely identify an observation. There are two types of keys:

1. Primary Key: Uniquely identifies an observation in its own table.
2. Foreign Key: Uniquely identifies an observation in another table.

## Mutating Joins
Mutating joins combine variables from two tables by matching observations based on their keys.

- inner_join(): Matches pairs of observations where keys are equal.
- left_join(): Keeps all observations in the left table.
- right_join(): Keeps all observations in the right table.
- full_join(): Keeps all observations in both tables.

Joins are visualized by matching rows based on key variables. 

## Filtering Joins
Filtering joins match observations and affect the observations, not the variables:

semi_join(): Keeps all observations in the left table that have a match in the right table.
anti_join(): Drops all observations in the left table that have a match in the right table.

## Set Operations
You can use set operations to compare data in tables.

Use intersect()if you have two tables with house data and want to find common entries,
Use union() to combine two tables without duplicates.
Use setdiff() to find rows present in one table but not in the other

When working with real data, identify the primary keys and make sure no primary key variables are missing 
Using anti_join() to check if all foreign key values in one table match primary key values in another.


# Chapter 11: Strings with stringr
It introduces string manipulation in R using stringr which is part of the stringr package.

## String Basics
Strings can be created using either single or double quotes. To include special characters like quotes or backslashes, use the escape character `\.` The writeLines() function shows the raw contents of a string.

## String Length
The str_length() function from stringr tells you the number of characters in a string:

```{r}
str_length(c("a", "R for data science", NA))

```
## Combining Strings
Use str_c() to combine two or more strings. The sep argument controls how they are separated, and collapse collapses a vector of strings into a single string:

```{r}
# Add details column, which is combines Location and BuildingType
house_sale <- house_sale %>%
  mutate(details = str_c(Location, BuildingType, sep = ", "))

head(house_sale$details)
```


## Subsetting Strings
The str_sub() function extracts parts of a string based on start and end positions. It can also be used to modify strings:

```{r}
x <- c("Apple", "Banana", "Pear")
str_sub(x, 1, 3)
```
## Locales
Functions like str_to_lower(), str_to_upper(), and str_to_title() change the case of text. We changed the case of the vectors in Building Type to small letters with the code below:

```{r}
house_sale <- house_sale %>%
    mutate(BuildingType = str_to_lower(BuildingType))

head(house_sale$BuildingType)
```

## Anchors
Anchors like ^ and $ match the start and end of a string, respectively. The word boundary \b matches the boundary between words.

## Character Classes and Alternatives
Special patterns match more than one character:

- `\d` matches any digit.
- `\s` matches any whitespace.
- [abc] matches a, b, or c.
- [^abc] matches anything except a, b, or c.

## Tools
Various stringr functions apply regexps to real problems:

1. str_detect(): Determines if a character vector matches a pattern.
2. str_extract(): Extracts the actual text of a match.
3. str_replace() and str_replace_all(): Replace matches with new values.
4. str_split(): Splits a string into pieces.
5. str_locate() and str_locate_all(): Give the starting and ending positions of each match.

# Chapter 12: Forecasts with Forecat
In R, factors are used to work with categorical variables, which have a fixed and known set of possible values. 
To work with factors, the forcats package is used, which provides tools for dealing with categorical variables. 

## Creating Factors
When recording variables like months, using strings can lead to issues such as typos and unhelpful sorting. These problems can be fixed by creating a factor with a predefined list of valid levels. For example:

```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")
month_levels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
y1 <- factor(x1, levels = month_levels)
y1
```
## Modifying Factor Order
Changing the order of factor levels in visualizations can make patterns easier to interpret. Functions like fct_reorder() and fct_relevel() are used to reorder levels based on specific criteria. For example, we can reorder Building type based on houseprice.

```{r}
# Reorder BuildingType based on average HousePrice
house_sale %>%
  ggplot(aes(x = HousePrice, y = fct_reorder(BuildingType, HousePrice))) + 
  geom_point() + 
  labs(title = "House Prices by Building Type", x = "House Price", y = "Building Type")


```

## Modifying Factor Levels
Changing the values of factor levels can clarify labels and collapse levels for high-level displays. The fct_recode() function allows users to recode factor levels. Basically renaming.

```{r}
#Converting EstateType from character to factor

house_sale <- house_sale %>%
  mutate(EstateType = as.factor(EstateType))

# Recode EstateType levels
house_sale <- house_sale %>% 
  mutate(EstateTypeNew = fct_recode(EstateType, 
                                 "Private-Builder" = "PrivateBuilder", 
                                 "Semi-Private" = "SemiPrivate",
                                 "Gvt-Build" = "GovernmentBuild",
                                 "Community" = "Society",
                                 "Other" = "Other"))

head(house_sale$EstateTypeNew)

```

To combine groups, multiple old levels can be assigned to the same new level. For collapsing many levels, fct_collapse() is useful. 

```{r}
#Converting EstateType from character to factor
house_sale <- house_sale %>%
  mutate(SaleType = as.factor(SaleType))

# Collapse SaleType levels into broader categories
house_sale <- house_sale %>% 
  mutate(SaleType2 = fct_collapse(SaleType, 
                                  "New" = c("NewHouse"), 
                                  "Old" = c("FifthResale", "FirstResale", "FourthResale", "SecondResale","ThirdResale")))

# Count new sale types
print(head(house_sale$SaleType2))

```
# Chapter 13: Date and times wit lubridate
This chapter explored working with dates and time in R. This will be done using the lubridate package.

There are three types of date/time data that refer to an instant in time:

1. A date. Tibbles print this as <date>.
2. A time within a day. Tibbles print this as <time>.
3. A datetime is a date plus a time: it uniquely identifies an instant in time (typically to the nearest second). Tibbles print this as <dttm>. Elsewhere in R these are called POSIXct.
To get the current date or datetime you can use today() or now():

```{r}
today()
now()
```

## From Strings
Date/time data often comes as strings. lubridate provides helpers that automatically work out the format once you specify the order of the component. For example:

```{r}
ymd("2017-01-31")         
mdy("January 31st, 2017") 
dmy("31-Jan-2017")        
```

To create a datetime, add an underscore and one or more of “h”, “m”, and “s” to the name of the parsing function:

```{r}
ymd_hms("2017-01-31 20:11:59") 
mdy_hm("01/31/2017 08:01")    
```
In some instances, different components of the datetime will be spread across multiple columns. To create a date/time from this sort of input, use make_date() for dates, or make_datetime() for datetimes.

You may want to switch between a datetime and a date. That’s the job of as_datetime() and as_date()

# Getting Components
You can pull out individual parts of the date with the accessor functions year(), month(), mday() (day of the month), yday() (day of the year), wday() (day of the week), hour(), minute(), and second().

```{r}
datetime <- ymd_hms("2016-07-08 12:34:56")
year(datetime)   
month(datetime)  
mday(datetime)   
yday(datetime)   
wday(datetime)   

```
For month() and wday() you can set label = TRUE to return the abbreviated name of the month or day of the week. 

## Time Spans
These classes represent time spans.

- Durations, which represent an exact number of seconds.
- Periods, which represent human units like weeks and months.
- Intervals, which represent a starting and ending point.

In R, when you subtract two dates, you get a difftime object.

Now for durations, durations come with a number of convenient constructors.

```{r}
dseconds(15)   
dminutes(10)   
dhours(c(12, 24)) 
ddays(0:5)    
dweeks(3)      
dyears(1) 

```

Periods are time spans but do not have a fixed length in seconds; instead, they work with “human” times, like days and months.

```{r}
one_pm <- ymd_hms("2016-03-12 13:00:00", tz = "America/New_York")
one_pm + days(1)

```
Period can also be constructed with friendly constructors.

```{r}
seconds(15)    
minutes(10)    
hours(c(12, 24)) 
days(7)       
months(1:6)    
weeks(3)       
years(1)      
```
## Timezones
One can find out what R thinks the current time zone is with Sys.timezone()
And see the complete list of all time zone names with OlsonNames()

```{r}
Sys.timezone()
length(OlsonNames())
head(OlsonNames())  
```


# Chapter 14: Pipes with magrittr
Pipes are used to express a sequence of multiple operations. Since the tidyverse library automatically loads dplyr and %>%, we won’t need to load magrittr explicitly.Applying the learnings of this chapter to my house_sale data:

## Basic Pipe (%>%): 
It streamlines multiple operations such as filtering and mutating.

```{r}
house_location<- house_sale %>%
  filter(Location == "Dhaula Kuan") %>%
  mutate(price_per_sqft = HousePrice / (StoreArea+BasementArea+LawnArea))

head(house_location %>% select(Location, price_per_sqft), 5)

```

## Tee Pipe (%T>%)
Use to print interim output while retaining the original data.

```{r}
house_location<- house_sale %>%
  filter(Location == "Dhaula Kuan") %>% 
  head(5)%T>%
  print() %>% 
  mutate(price_per_sqft = HousePrice / (StoreArea+BasementArea+LawnArea))

```

## Explode Pipe (%$%): 
Apply calculations directly between specific columns.

```{r}
house_sale %$%
  cor(HousePrice, BasementArea)  # Finds the correlation between HousePrice and BasementArea
```
## Assignment Pipe (%<>%): 
Mutate house_sale directly without reassigning each time.

```{r}
house_location %<>% 
  mutate(price_per_sqft2 = HousePrice / BasementArea)

print(head(house_location$price_per_sqft2))
```
# Chapter 15: Functions
Functions in R help automate repetitive tasks, making your code cleaner and easier to maintain.


## Writing functions examples
Below, we are trying to calculate squares of numbers and returning result as a string.

```{r}
square_number <- function(x, as_string = FALSE) {
  result <- x^2
  if (as_string) {
    return(paste("The square of", x, "is", result))
  }
  return(result)
}
print(square_number(10,TRUE))

```
We can also create a function to check if a number is positive or negative.

```{r}
check_number <- function(x) {
  if (x > 0) {
    return("Positive")
  } else if (x < 0) {
    return("Negative")
  } else {
    return("Zero")
  }
}

check_number(10)
check_number(0)
check_number(-50)

```
## Dot-Dot-Dot (...) for Flexible Functions

```{r}
combine_strings <- function(...) {
  return(paste(..., collapse = " "))
}

# Using the function
print(combine_strings("We", "can't", "overlook", "the", "effect", "of", "game", "theory"))  # 

```

## Lazy Evaluation
this is a function that only evaluates its arguments when needed.
```{r}
lazy_evaluation_example <- function(x) {
  if (x == 0) {
    return("Zero is not allowed")
  }
  return(1/x)
}

# Using the function
print(lazy_evaluation_example(0))  
print(lazy_evaluation_example(20))

```

## Returning Values and Early Returns
The below function that sums two numbers but returns early if one of them is zero.

```{r}
sum_numbers <- function(a, b) {
  if (a == 0 || b == 0) {
    return(0)
  }
  return(a + b)
}

# Using the function
print(sum_numbers(5, 3))
print(sum_numbers(0, 3))

```
